Type Classes for Mathematics in Type Theory

### 2010-06-28 jar.tex
Replace some "i.e."'s. I'm less eager to replace "e.g."'s.

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g. rewriting).

@@

Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (i.e. proof terms).

@@

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (i.e.\ without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.\marginpar{something about opacity?}

@@

A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful). Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g. coercion pullbacks~\cite{Hints}).

@@

  Class Functor `(Fmap): Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎ g) = fmap f ◎ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation (i.e. ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.

>>>>

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting).

@@

Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).

@@

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.\marginpar{something about opacity?}

@@

A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful). Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g.\ coercion pullbacks~\cite{Hints}).

@@

  Class Functor `(Fmap): Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎ g) = fmap f ◎ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.
